---
title: "Logistic regression and Poisson regression"
author: "Anders Christiansen Sørby, Edvard Hove"
date: "October 4, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Logistic regression

In this part we model the probability of a successful ascent of a mountain using binary regression with a logit link.
Let $y_i$ be the number of successful ascents, and let $n_i$ be the total number of attempts (sum of successful and failed) of the $i$th mountain.
The GLM model is then,

1. Model for response: $Y_i \sim \text{Bin}(n_i, \pi_i) \quad \text{for} \  i = 1,\ldots,113$
2. Linear predictor: $\eta_i=x^T_i\beta$
3. Link function: $\eta_i = \ln \left(\frac{\pi_i}{1-\pi_i}\right)$

where $\mathbf{x}_i$ is $p$ dimensional column vector of covariates for observation $i$, and $\boldsymbol\beta$ is the vector of regression parameters.

## a)

Since we assume independent pairs of $(y_i,\mathbf{x}_i)$, the likelihood as a function can then be written as

\begin{align}
L(\beta) = \prod_{i=1}^n\pi_i^{y_i}(1-\pi_i)^{1-y_i}
\end{align}

the log-likelihood as a function of $\beta$ is then found by taking the logarithm and using the link function to replace $pi_i$.

\begin{align}
\ell(\beta) &= \sum_{i=1}^n y_i \ln \pi_i + (1-y_i) \ln (1-\pi_i) \\
&= \sum_{i=1}^n y_i \ln \left(\frac{\pi_i}{1-\pi_i}\right) + \ln(1-\pi_i) \\
&= \sum_{i=1}^n y_i \mathbf{x}_i \beta - \ln(1+e^{\mathbf{x}_i \beta})
\end{align}

The maximum likelihood estimate of $\beta$ is then found by maximizing $\ell(\beta)$.
For some models this can be expressed on closed form by solving 

\begin{align}
s(\hat \beta) = \frac{\partial \ell(\hat \beta)}{\partial \beta} = 0
\end{align}

In this case the equations above are non-linear, and the maximum is found using numerical methods like Newton-Raphson or Fisher scoring.

## b)

```{r}
filepath <- "https://www.math.ntnu.no/emner/TMA4315/2018h/mountains"
mount <- read.table(file = filepath, header = TRUE, 
                    col.names = c("height", "prominence", "fail", "success"))
```

```{r}
model1 <- glm(cbind(success, fail) ~ height + prominence, data = mount, family = "binomial")
summary(model1)
```




# Part 2: Poisson regression


```{r tippeligaen}
filepath <- "https://www.math.ntnu.no/emner/TMA4315/2018h/eliteserien2018"
tippeliga <- read.table(file = filepath, header = TRUE,
                        colClasses = c("character", "character", "numeric", "numeric"))
```

```{r myglm}
source('myglm.R')
```

## a)

Indepence between the goals of the home and the away team is questionable. 



# Source code
```{r source, code=readLines('myglm.R')}

```